---
title: "Bayesian updating of EXCEL results - Figure 2"
author: "J Brophy"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    highlight: tango
    theme: cerulean
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
link-citations: yes
linkcolor: red
urlcolor: blue
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=3)
```

## Load data

Data set was created in Figure1.Rmd file. This file takes the EXCEL data and updates it with an informative prior based on similar completed RCTs.

```{r results='asis', warning=FALSE, message=FALSE}

library(tidyverse)
library(knitr)
library(kableExtra)
library(ggthemes)
library(metafor)
library(meta)

# Load dataset created and saved in Figure1.Rmd
temp <- read.csv("data1.csv", header = TRUE)
prior <- temp[1:3,]
```

## Primary outcome
The first analysis examines the EXCEL primary outcome - death, non-fatal MI, and stroke - from both the prior data and the new (EXCEL) data. 



### Function for congugate normal analyses

Given the large sample sizes the distributions can be approximated as normal.    
Here is a function that takes the prior and likelihood risk differences, both expressed as a mean and variance, and combines them in a conjugate normal analysis to get the posterior distributions.

```{r}
post.normal.mean <- function(data.mean, data.var, prior.mean, prior.var)
{
####################################################################
#  R function for Bayesian analysis of normal mean, variance known #
#  Parameters included are:                                        #
#                                                                  #
#  Inputs:                                                         #
#                                                                  #
#   x = vector of data                                             #
#   prior.mean = prior mean                                        #
#   prior.var  = prior variance                                    #
#   data.var   = assumed known variance of data                    #
#                                                                  #
#  Outputs:                                                        #
#                                                                  #
#   post.mean = posterior mean                                     #
#   post.var  = posterior variance                                 #
#                                                                  #
####################################################################

post.mean.numerator <- prior.mean/prior.var + data.mean/data.var
post.mean.denominator <- 1/prior.var + 1/data.var
post.mean <-  post.mean.numerator/post.mean.denominator
post.var <- (1/(1/prior.var + 1/data.var))
a <- "Post mean = "
b <- "Post Var = "
c <- "Post SD = "
cat(a, post.mean, ",", b, post.var, ",", c, sqrt(post.var), "\n" )
newlist <- list(post.mean, post.var, sqrt(post.var))
return(newlist)
}

```


### Primary outcome using prior information    

#### Random effects model of previous studies

We now use estimates of effect size and the variance of the 3 previous studies obtained under a random effects model which includes both the within and between study variances.     
We will combine this prior information with the EXCEL data using the congugate normal function defined previously. 

```{r}
# prior primary outcome, random effects model for the risk differences
prior <- temp[1:3,]
RD_re <- rma(ai=Pe, n1i= Ne, ci=Pc, n2i=Nc, data=prior, measure="RD",
           slab=paste(Study), method="REML")

# from above rma function 
prior.diff.prim.mean <-RD_re$beta
prior.diff.prim.sd <- RD_re$se

```
#### Combing EXCEL with prior information

```{r}
# same procedure for the EXCEL data (likelihood)
like.e.prim.mean <-temp[4,5]/temp[4,9] # like PCI (experimental) for 1o outcome
like.c.prim.mean <-temp[4,4]/temp[4,8]
like.diff.prim.mean <- like.e.prim.mean- like.c.prim.mean

like.e.prim.sd <-sqrt(like.e.prim.mean * (1-like.e.prim.mean) / temp[4,9])
like.c.prim.sd <-sqrt(like.c.prim.mean * (1-like.c.prim.mean) /temp[4,8])
like.diff.prim.sd <-sqrt(like.e.prim.sd^2 + like.c.prim.sd^2)

# EXCEL primary outcome + random effects model of previous studies combined with normal conjugacy
data.mean <- like.diff.prim.mean # EXCEL data
data.var <- like.diff.prim.sd^2

prior.mean <- RD_re$beta  # rma estimates
prior.var <- RD_re$se^2 

post.primary <- post.normal.mean(data.mean, data.var, prior.mean, prior.var)
post.primary
```

#### Triplot for primary outcome Figure 2a

```{r}
df <- data.frame(100*rbind(c(prior.diff.prim.mean, prior.diff.prim.sd), c(like.diff.prim.mean, like.diff.prim.sd), c(post.primary[[1]], post.primary[[3]])))
df$type <- c("Prior", "EXCEL", "Posterior")
df <- df %>%
  rename(mean = X1, se = X2)

fig2a <- ggplot(data.frame(x = c(-4, 10)), aes(x = x)) +
        stat_function(fun = dnorm, args = list(post.primary[[1]]*100, post.primary[[3]]*100), colour = "deeppink") +
        stat_function(fun = dnorm, args = list(0.0181*100, 0.0275*100), colour = "blue") +
        stat_function(fun = dnorm, args = list(like.diff.prim.mean*100, like.diff.prim.sd*100), colour = "dark green") +
        ggtitle("Triplot of primary outcome \n(prior, likelihood and posterior PDFs)") +
        xlab ("Primary outcome difference (PCI - CABG)") +
        ylab ("Density") +
        annotate("text", label = "Blue line \n prior information \n (SYNTAX, NOBLE \n& PRECOMBAT) ", x = -2, y = .1, color = "black") +
        annotate("text", label = "Green line \n likelihood from EXCEL ", x = 6.5, y = .11, color = "black") + 
        annotate("text", label = "Red line posterior (all data) \n probability > 0 = 96% (yellow + grey) \nprobability > 1 = 86% (grey) ", x = 5.5, y = .22, color = "black") +
        theme_economist() +
        stat_function(fun = dnorm, args = list(post.primary[[1]]*100, post.primary[[3]]*100), 
                xlim = c(1,8), geom = "area", alpha = 0.2) +
        stat_function(fun = dnorm, args = list(post.primary[[1]]*100, post.primary[[3]]*100), 
                xlim = c(0,1), geom = "area", alpha = 0.2, fill = "yellow") 
fig2a
paste(" Combined posterior mean of EXCEL data and prior studies while accounting for betweeen study variation = ", round(post.primary[[1]]*100,2), " (", round(post.primary[[3]]*100, 2), ") more outcomes with PCI")

paste("95% CrI", round(post.primary[[1]]*100 - 1.96 * post.primary[[3]]*100,2), " - ", round(post.primary[[1]]*100 + 1.96 * post.primary[[3]]*100,2))  

paste(" Probability that primary outcome PCI risk > CABG risk")
1-pnorm(0, post.primary[[1]], post.primary[[3]])
paste(" Probability that primary outcome PCI risk > CABG risk by at least 1 / 100 treated")
1-pnorm(.01, post.primary[[1]], post.primary[[3]])
```


### Secondary (Mortality) Outcome

```{r}
# likelihood EXCEL mortality

like.e.mort.mean <-temp[4,3]/temp[4,9] # like PCI (experimental) for 1o outcome
like.c.mort.mean <-temp[4,2]/temp[4,8]
like.diff.mort.mean <- like.e.mort.mean- like.c.mort.mean

like.e.mort.sd <-sqrt(like.e.mort.mean * (1-like.e.mort.mean) / temp[4,9])
like.c.mort.sd <-sqrt(like.c.mort.mean * (1-like.c.mort.mean) /temp[4,8])
like.diff.mort.sd <-sqrt(like.e.mort.sd^2 + like.c.mort.sd^2)

# prior
# from rma function program
prior <- temp[1:3,]
RD_re <- rma(ai=Me, n1i= Ne, ci=Mc, n2i=Nc, data=prior, measure="RD",
           slab=paste(Study), method="REML")

# EXCEL mortality outcome + random effects model of previous studies combined with normal conjugacy
data.mean <- like.diff.mort.mean # EXCEL data
data.var <- like.diff.mort.sd^2
prior.mean <- RD_re$beta  # rma estimates
prior.var <- RD_re$se^2 
post.primary <- post.normal.mean(data.mean, data.var, prior.mean, prior.var)
post.primary

paste("Posterior probability of PCI mortality > CABG  and > by 1 in 100 treated")
1-pnorm(0, post.primary[[1]], post.primary[[3]])
1-pnorm(.01, post.primary[[1]], post.primary[[3]])

```

#### Triplot for mortality outcome Figure 2b

```{r}
df <- data.frame(100*rbind(c(RD_re$beta, RD_re$se), c(like.diff.mort.mean, like.diff.mort.sd), c(post.primary[[1]], post.primary[[3]])))
df$type <- c("Prior", "EXCEL", "Posterior")
df <- df %>%
  rename(mean = X1, se = X2)

fig2b <- ggplot(data.frame(x = c(-4, 8)), aes(x = x)) +
        stat_function(fun = dnorm, args = list(post.primary[[1]]*100, post.primary[[3]]*100), colour = "deeppink") +
        stat_function(fun = dnorm, args = list(RD_re$beta*100, RD_re$se*100), colour = "blue") +
        stat_function(fun = dnorm, args = list(like.diff.mort.mean*100, like.diff.mort.sd*100), colour = "dark green") +
        ggtitle("Triplot of mortality outcome \n(prior, likelihood and posterior PDFs)") +
        xlab ("Mortality difference (PCI - CABG)") +
        ylab ("Density") +
        annotate("text", label = "Blue line \n prior information \n (SYNTAX, NOBLE \n& PRECOMBAT) ", x = -2, y = .1, color = "black") +
        annotate("text", label = "Green line \n likelihood from EXCEL ", x = 5, y = .11, color = "black") + 
        annotate("text", label = "Red line posterior (all data) \n probability > 0 = 85% (yellow + grey) \nprobability > 1 = 47% (grey) ", x = 4, y = .35, color = "black") +
        theme_economist() +
        stat_function(fun = dnorm, args = list(post.primary[[1]]*100, post.primary[[3]]*100), 
                xlim = c(1,8), geom = "area", alpha = 0.2) +
        stat_function(fun = dnorm, args = list(post.primary[[1]]*100, post.primary[[3]]*100), 
                xlim = c(0,1), geom = "area", alpha = 0.2, fill = "yellow") 
fig2b
paste(" Combined posterior mean of EXCEL data and prior studies while accounting for betweeen study variation = ", round(post.primary[[1]]*100,2), " (", round(post.primary[[3]]*100, 2), ") more outcomes with PCI")
post.primary[[1]]*100 # posterior mean
post.primary[[3]]*100 # posterior SE
post.primary[[1]]*100 + c(-1,1) * 1.96 * post.primary[[3]]*100  # posterior 95% CI

```

### Secondary (MACCE) outcome

```{r}
# likelihood EXCEL mortality

like.e.macce.mean <-temp[4,7]/temp[4,9] # like PCI (experimental) for 1o outcome
like.c.macce.mean <-temp[4,6]/temp[4,8]
like.diff.macce.mean <- like.e.macce.mean- like.c.macce.mean

like.e.macce.sd <-sqrt(like.e.macce.mean * (1-like.e.macce.mean) / temp[4,9])
like.c.macce.sd <-sqrt(like.c.macce.mean * (1-like.c.macce.mean) /temp[4,8])
like.diff.macce.sd <-sqrt(like.e.macce.sd^2 + like.c.macce.sd^2)

# prior
# from rma function program
prior <- temp[1:3,]
RD_re <- rma(ai=Re, n1i= Ne, ci=Rc, n2i=Nc, data=prior, measure="RD",
           slab=paste(Study), method="REML")

# posterior
# EXCEL primary outcome + random effects pooling of previous studies
data.mean <- like.diff.macce.mean # EXCEL data
data.var <- like.diff.macce.sd^2
prior.mean <- RD_re$beta # rma estimates
prior.var <- RD_re$se^2
post.primary <- post.normal.mean(data.mean, data.var, prior.mean, prior.var)
post.primary
```

#### Triplot for MACCE outcome Figure 2c


```{r}
df <- data.frame(100*rbind(c(RD_re$beta, RD_re$se), c(like.diff.macce.mean, like.diff.macce.sd), c(post.primary[[1]], post.primary[[3]])))
df$type <- c("Prior", "EXCEL", "Posterior")
df <- df %>%
  rename(mean = X1, se = X2)

fig2c <- ggplot(data.frame(x = c(0, 12)), aes(x = x)) +
        stat_function(fun = dnorm, args = list(post.primary[[1]]*100, post.primary[[3]]*100), colour = "deeppink") +
        stat_function(fun = dnorm, args = list(RD_re$beta*100, RD_re$se*100), colour = "blue") +
        stat_function(fun = dnorm, args = list(like.diff.macce.mean*100, like.diff.macce.sd*100), colour = "dark green") +
        ggtitle("Triplot of MACCE outcome \n(prior, likelihood and posterior PDFs)") +
        xlab ("MACCE difference (PCI - CABG)") +
        ylab ("Density") +
        annotate("text", label = "Blue line \n prior information \n (SYNTAX, NOBLE \n& PRECOMBAT) ", x = 2, y = .08, color = "black") +
        annotate("text", label = "Green line \n likelihood from EXCEL ", x = 10.5, y = .11, color = "black") + 
        annotate("text", label = "Red line posterior (all data) \n probability > 0 = 99.9% (yellow + grey) \nprobability > 5 = 90% (grey) ", x = 2.8, y = .21, color = "black") +
        theme_economist() +
        stat_function(fun = dnorm, args = list(post.primary[[1]]*100, post.primary[[3]]*100), 
                xlim = c(5,12), geom = "area", alpha = 0.2) +
        stat_function(fun = dnorm, args = list(post.primary[[1]]*100, post.primary[[3]]*100), 
                xlim = c(0,5), geom = "area", alpha = 0.2, fill = "yellow") 
fig2c
post.primary[[1]]*100 # posterior mean
post.primary[[3]]*100 # posterior SE
paste("95% CrI", round(post.primary[[1]]*100 - 1.96 * post.primary[[3]]*100,2), " - ", round(post.primary[[1]]*100 + 1.96 * post.primary[[3]]*100,2))  
paste(" Probability > 0 - 5 extra events ")
1-pnorm(0, post.primary[[1]], post.primary[[3]])
1-pnorm(.01, post.primary[[1]], post.primary[[3]])
1-pnorm(.02, post.primary[[1]], post.primary[[3]])
1-pnorm(.03, post.primary[[1]], post.primary[[3]])
1-pnorm(.04, post.primary[[1]], post.primary[[3]])
1-pnorm(.05, post.primary[[1]], post.primary[[3]])
```

## Combining graphs to pdf

```{r}
pdf("fig2.pdf")
plot(fig2a)
plot(fig2b)
plot(fig2c)
dev.off()
```

